---
title: "Cleaning OMB FOIA Data"
author: "Ryan Harrington"
output: html_document
---

# Overview

The Freedom of Information Act (FOIA) allows constituents to gain access to the data that their government creates. Gaining access to this data comes through a process called a "FOIA request". Making a FOIA request does not guarantee that the request will be fulfilled. There are many reasons that a request may be denied - from the data being unavailable to the data falling under executive privilege.

In Delaware, every government agency is required by law to keep a FOIA log. This log is a historical record of all FOIA requests that have been made to that agency. In an effort to better understand what data was commonly being FOIA requested, Open Data Delaware made a FOIA request to as many agencies as possible for their FOIA logs. Most of these FOIA-log-FOIA-requests were made on February 9, 2019 by David Ginzberg.

As the FOIA-log-FOIA-requests have been fulfilled, we have the opportunity to analyze each agency's FOIA logs to better understand their data. To begin this analysis, we look at the Office of Management and Budget's FOIA logs.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}

# For file management
# install.packages("here")
library(here)

# For data manipulation, functional programming, and graphing
# install.packages("tidyverse")
library(tidyverse)

# For reading in data from Excel
# install.packages("readxl")
library(readxl)

```

# Import Data

```{r}

# Read in the `OMB FOIA TRACKING 032719.xlsx` file from the `Data` folder in the
# project directory and then store it in an object called foia_log_OMB
foia_log_OMB <- read_excel(here("Data", "OMB FOIA TRACKING 032719.xlsx"))

```

# Data Cleaning

Column names are unwieldy for two reasons:

1. They are long. This makes them extremely descriptive, but annoying to work with.
2. They utilize non-alphanumeric characters (for example, *, -, and ยง).

Removing these issues helps to make our work easier later in the analysis.

Throughout the analysis you will see the piping operator, `%>%`. If you are not familiar with the piping operator, you can read it as "and then". Review the comments within the code to get a better understanding of how it is being used.

```{r column names}

# First, we can remove any characters that are not A-Z, a-z, or 0-9.
# Next, we replace all spaces with underscores.
colnames(foia_log_OMB) <-
  colnames(foia_log_OMB) %>%                  # Get a list of all column names in the object foia_log_OMB AND THEN
  str_replace_all(pattern = "[^A-Za-z0-9 ]",  # Replace any characters that are not A-Z, a-z, or 0-9
                  replacement = "") %>%       # with nothing. AND THEN
  str_replace_all(pattern = " ",              # Replace any spaces
                  replacement = "_")          # with underscores.


# Some column names will need to be shortened manually.
# The rename function allows us to assign column names with new names (in quotes).
foia_log_OMB <-
  foia_log_OMB %>%                            # Use the data frame called foia_log_OMB AND THEN rename columns such as:
  rename("Date_Received" = Date_Written_Request_Received_By_Agency,
         "Date_Assigned" = Assigned_Date,
         "Division" = OMB_Division,
         "Date_Review" = Agency_review_date,
         "Is_Noncustodial_Record" = Does_the_request_seek_noncustodial_records_from_DTI_OMB_or_any,
         "Noncustodial_Provided_Other_Agency" = If_noncustodial_records_were_provided_by_another_agency_provid,
         "Date_Response" = Date_of_Agency_Response_to_Requesting_Party_Under_34_of_Policy,
         "Date_Estimate_Sent" = Estimate_Sent,
         "All_Requested_Records_Sent" = Were_all_requested_records_provided_on__date_of_Agencys_respons,
         "Date_Requested_Records_Not_Provided" = If_all_requested_records_were_not_provided_as_of_date_of_Agency,
         "Copying_Fees" = Copying_Fees_Note1st_20_pages_free)

```

Each column should be checked individually to identify any cleaning opportunities. Start by looking at the structure of the dataset.

```{r structure}

str(foia_log_OMB)

```

We see a mix of `POSIXct`, `chr`, `num`, and `logical` columns. `POSIXct` is a type of date field. `chr` is a character field. `num` is a numeric field. `logical` is a logical or `TRUE`/`FALSE` field. Quickly reviewing each field, we see that the fields look like they largely make sense. However, `Date_Requested_Records_Not_Provided` seems to be listed as a character field, when it may be more appropriate to list it as a date field. This is something we may have to deal with later.

Next, we should get a summary of the columns. This is particularly useful for date and numeric columns, though we do get some information about the character fields as well. For date and numeric fields we will get the following information:

* `Min.` - the minimum value
* `1st Qu.` - the first quartile value
* `Median` - the median value
* `Mean` - the average value
* `3rd Qu.` - the third quartile value
* `Max.` - the maximum value
* `NA's` - the number of `NULL` values (if any)

```{r summary}

summary(foia_log_OMB)

```

Looking at this information, we notice a few things that are interesting immediately:

1. The `Date_Received` field is never `NULL`. This inherently makes sense.
2. The `Date_Assigned` field is never `NULL`. This tells us that every FOIA request that is made to the Office of Management and Budget is assigned to a person.
3. The `Date_Review` field is `NULL` 11 times. Apparently, 11 FOIA requests that were made did not require review.
4. The `Date_Response` field is `NULL` 5 times. There have been 5 times where there was no response when a request was made.
5. The `Date_Estimate_Sent` field is `NULL` 447 times. Most times that a FOIA request is made, there is no need to send a cost estimate.
6. The `Copying_Fees` field is `NULL` 427 times. This aligns with the `Date_Estimate_Sent` field in that most times a FOIA request is made, there is no need to send a cost estimate.
7. The `Administrative_Fees` field is `NULL` 499 times. Similarly, this aligns with the `Date_Estimate_Sent` field. It is interesting to note that while it is logical for all of these fields to have `NULL` values, they are not `NULL` the same number of times. We should determine if there is a relationship between the three fields.
8. The `Date_of_Final_Disposition` field is `NULL` 11 times. Apparently, 11 FOIA requests that were made did not require this to occur. It is interesting to note that the `Date_Review` field was also `NULL` 11 times. Perhaps whenever one of these fields is `NULL`, then the other is as well.
9. The `Documents` field is `NULL` 523 times. This is the same number as how many rows there are in the full dataset. Therefore, we can discard the `Documents` field.

```{r}

# Remove the Documents field
foia_log_OMB <- 
  foia_log_OMB %>%                            # Use the data frame called foia_log_OMB AND THEN
  select(-Documents)                          # select all columns EXCEPT for Documents

```

Now, we explore the number of unique values in each column. Some of the columns have a large number of unique values. Others have much less. We might want to explore some of the columns with relatively few unique values to get a better understanding of the counts of each unique value. In doing so, we may find that some additional data cleaning is needed to ensure consistency within the dataset.

```{r unique values}

# Find the number of unique values for each column in the dataset
(foia_log_OMB_unique_values <-
   foia_log_OMB %>%                                         # Use the data frame called foia_log_OMB AND THEN
   purrr::map_df(function(x) length(unique(x))) %>%         # find the number of unique values in each field AND THEN
   t() %>%                                                  # transpose the created list AND THEN
   as.data.frame() %>%                                      # convert it to a data frame AND THEN
   rownames_to_column() %>%                                 # convert the row names to a column AND THEN
   rename("Column_Name" = rowname, "Unique_Values" = V1))   # rename each column to something more useful

# Note that the previous call is wrapped fully in parentheses. In an RMarkdown
# file, when we wrap a statement in parentheses, it is automatically included
# as an output. In this case, we are able to both store the data into a new
# object and include it in the output. It's a nice little trick.

# Create a list of columns for further exploration
foia_log_OMB_unique_values_selected <-
  foia_log_OMB_unique_values %>%                            # Use the data frame called foia_log_OMB_small_unique_values AND THEN
  filter(Unique_Values < 30) %>%                            # filter the data frame such that we only include Unique_Values less than 30 AND THEN
  select(Column_Name) %>%                                   # select the column "Column_Name" AND THEN
  unlist() %>%                                              # unlist the dataframe AND THEN
  unname()                                                  # unname the dataframe

```

Of the different fields with a low number of unique values, we'll highlight the `Assigned_to` field. This is a great example of why data cleaning is so important.

```{r}

foia_log_OMB %>%                              # Use the data frame called foia_log_OMB AND THEN
  count(Assigned_to)                          # count each unique value in the field `Assigned_to`

```

When observing the counts of each unique value in the `Assigned_to` field, we can see that there should _really_ be 3 values - Bert, Jessica, and `NULL`. However, there are some clear data cleanliness issues with how Bert is spelled. This can easily be fixed.

```{r}

foia_log_OMB <-
  foia_log_OMB %>%                                                # Use the data frame called foia_log_OMB AND THEN
  mutate(Assigned_to = str_remove_all(string = Assigned_to,       # remove any characters in `Assigned_to`
                                      pattern = "[^A-Za-z0-9]"))  # that are not A-Z, a-z, or 0-9

```

After doing this, we can see that our data now correctly reflects that there are 3 values in the `Assigned_to` field - Bert, Jessica, and `NULL`.

```{r}

foia_log_OMB %>%                              # Use the data frame called foia_log_OMB AND THEN
  count(Assigned_to)                          # count each unique value in the field `Assigned_to`

```

We will need to repeat this procees for each field in order to ensure that our data is as clean as possible. Moving through the list of selected fields to explore, next is `Division`.

```{r}

foia_log_OMB %>% 
  count(Division)

```

It is very clear that some of these unique values should actually be represented as the same value. An obvious example of this are all of the variations of "Director's Office". We could also do the same for "Benefits", "Pension", and "PHRST". Before doing this cleaning, there are 23 unique values. After consolidating this, we'll be able to see this number be reduced. Outside of the few instances that we've listed, there are several other values that may be able to collapse into each other. For example "HR" and "HRM" could potentially represent the same division. However, that is less obvious, and so we should consult with the owner of the dataset for clarity.

```{r}

foia_log_OMB <-
  foia_log_OMB %>% 
  mutate(Division = if_else(str_detect(string = Division, pattern = "Dir"),
                            true = "Director's Office",
                            false = Division),
         Division = if_else(str_detect(string = Division, pattern = "Benefits"),
                            true = "Benefits Office",
                            false = Division),
         Division = if_else(str_detect(string = Division, pattern = "Pension"),
                            true = "Pensions",
                            false = Division),
         Division = if_else(str_detect(string = Division, pattern = "PHRST"),
                            true = "PHRST",
                            false = Division))

foia_log_OMB %>% 
  count(Division)

```

Exploration of the `Is_Noncustodial_Record` field shows that it can also be collapsed.

```{r}

foia_log_OMB %>% 
  count(Is_Noncustodial_Record)

```

The values tend to either be "Yes" or "No". We can see that we are running into issues with case sensitivity. Also, one record has additional explanation in it. For the purposes of this analysis, we will ignore this. However, it could be important to consider in a different analysis. Also worth noting, is that it may be more convenient for future analyses to code "Yes" values as 1's and "No" values as 0's. For now, we choose to not do this.

```{r}

foia_log_OMB <-
  foia_log_OMB %>% 
  mutate(Is_Noncustodial_Record = case_when(str_detect(string = Is_Noncustodial_Record,
                                                       pattern = "[Nn][Oo]") ~ "No",
                                            str_detect(string = Is_Noncustodial_Record,
                                                       pattern = "[Yy][Ee][Ss]") ~ "Yes"))

foia_log_OMB %>% 
  count(Is_Noncustodial_Record)

```

The `Noncustodial_Provided_Other_Agency` field has 14 unique values, each of which only appears one time.

```{r}

foia_log_OMB %>%
  count(Noncustodial_Provided_Other_Agency)

```

There are several changes that could occur here. We see some instances where the same agency is listed (in particular, DHSS), but it is listed differently each time. There are also times where there are values that should be coded as `NULL`, but instead receive a value ("N/A" and "No"). All of these issues must be fixed. Noting how sparse this particular field is, you may decide that doing this is not worth your time and may choose to exclude it from your analysis.

```{r}

foia_log_OMB <-
  foia_log_OMB %>%
  
  # Replace any field that includes DHSS with DHSS
  mutate(Noncustodial_Provided_Other_Agency = if_else(str_detect(string = Noncustodial_Provided_Other_Agency,
                                                                 pattern = "DHSS"),
                                                      true = "DHSS",
                                                      false = Noncustodial_Provided_Other_Agency),
         
         # Replace any field that includes Angela with DOF (she must be the representative for the Department of Finance)
         Noncustodial_Provided_Other_Agency = if_else(str_detect(string = Noncustodial_Provided_Other_Agency,
                                                                 pattern = "Angela"),
                                                      true = "DOF",
                                                      false = Noncustodial_Provided_Other_Agency),
         
         # Replace any field that includes N/A or No with an actual NA.
         # Note - the base R version of ifelse must be used because the vectorised dplyr version maintains type
         # which throws an error. Finding a better solution to this would be ideal.
         Noncustodial_Provided_Other_Agency = ifelse(str_detect(string = Noncustodial_Provided_Other_Agency,
                                                                 pattern = "N/A|No"),
                                                      yes = NA,
                                                      no = Noncustodial_Provided_Other_Agency),
         
         # Remove "Wendy Brown, " from all fields
         Noncustodial_Provided_Other_Agency = str_replace_all(string = Noncustodial_Provided_Other_Agency,
                                                              pattern = "Wendy Brown, ",
                                                              replacement = ""),
         
         # Remove "Referred to " from all fields
         Noncustodial_Provided_Other_Agency = str_replace_all(string = Noncustodial_Provided_Other_Agency,
                                                              pattern = "Referred to ",
                                                              replacement = ""),
         
         # Replace DSCYTF with the more commonly listed DSCYF
         Noncustodial_Provided_Other_Agency = str_replace_all(string = Noncustodial_Provided_Other_Agency,
                                                              pattern = "DSCYTF",
                                                              replacement = "DSCYF"))

foia_log_OMB %>% 
  count(Noncustodial_Provided_Other_Agency)

```

We see the same issue with the `Internally_Reviewed_By` field that we did with the `Assigned_to` field.

```{r}

foia_log_OMB %>% 
  count(Internally_Reviewed_By)

```

This means that we can fix it using the same technique.

```{r}

foia_log_OMB <-
  foia_log_OMB %>%
  mutate(Internally_Reviewed_By = str_remove_all(string = Internally_Reviewed_By,
                                                 pattern = "[^A-Za-z0-9]"))

foia_log_OMB %>% 
  count(Internally_Reviewed_By)

```

For the `All_Requested_Records_Sent` field, we see similar issues to the `Is_Noncustodial_Record` field. However, this time there is even further detail provided.

```{r}

foia_log_OMB %>% 
  count(All_Requested_Records_Sent)

```

The best strategy for cleaning this data may be to separate the "Yes" or "No" answers from the detail included in the answer. This means that we need to create an additional column to handle the detail.

```{r}

foia_log_OMB <- 
  foia_log_OMB %>% 
  mutate(All_Requested_Records_Sent = str_replace(string = All_Requested_Records_Sent,
                                                  pattern = "^[Nn][Oo]",
                                                  replacement = "No"),
         All_Requested_Records_Sent = str_replace(string = All_Requested_Records_Sent,
                                                  pattern = "^[Yy][Ee][Ss]|^[Yy][Rr][Dd]",
                                                  replacement = "Yes"),
         All_Requested_Records_Sent_Flag = case_when(str_detect(string = All_Requested_Records_Sent,
                                                                pattern = "^[Yy][Ee][Ss]") ~ "Yes",
                                                     str_detect(string = All_Requested_Records_Sent,
                                                                pattern = "^[Nn][Oo]") ~ "No",
                                                     str_detect(string = All_Requested_Records_Sent,
                                                                pattern = "^[Pp][Aa][Rr][Tt][Ii][Aa][Ll]") ~ "Partial",
                                                     str_detect(string = All_Requested_Records_Sent,
                                                                pattern = "^(?!.*No|Yes|Partial).*$") ~ "Other"),
         All_Requested_Records_Sent_Detail = str_replace(string = All_Requested_Records_Sent,
                                                         pattern = "^[Yy][Ee][Ss]|^[Nn][Oo]|^[Pp][Aa][Rr][Tt][Ii][Aa][Ll]",
                                                         replacement = ""),
         All_Requested_Records_Sent_Detail = str_replace(string = All_Requested_Records_Sent_Detail,
                                                         pattern = "^, ",
                                                         replacement = ""))

foia_log_OMB %>% 
  count(All_Requested_Records_Sent_Flag, All_Requested_Records_Sent_Detail)

```

Looking into the `Date_Requested_Records_Not_Provided` field shows us that this field has a number of problems.

```{r}

foia_log_OMB %>% 
  count(Date_Requested_Records_Not_Provided)

```

The primary issue is that it seems to be holding two types of information - dates and notes. Beyond this, the field is primarily empty, with only 11 non-missing values. Due to all of these issues, it makes more sense to simply avoid using this data than it does to try making use of it in the context of this analysis.

Instead, we can concentrate on our last field - `Administrative_Fees`.

```{r}

foia_log_OMB %>% 
  count(Administrative_Fees)

```

After inspection, it appears that there is no data cleaning that needs to occur with the `Administrative_Fees` field at this time.

Last, we check the `Current_Status` field. While there are a large number of unique values (82), there are appears to be a great deal of overlap between them. We can reduce the number of unique values here by grouping the different values of `Current Status` together.

```{r}

foia_log_OMB %>% 
  count(Current_Status) %>% 
  arrange(-n)

```

We can split the `Current Status` into two new fields - `Current_Status_Overall` and `Current_Status_Detail`.

```{r}

foia_log_OMB <-
  foia_log_OMB %>%
  mutate(Current_Status_Overall = str_extract(string = Current_Status,
                                              pattern = "^Closed|^Open|^Inactive"),
         Current_Status_Detail = case_when(str_detect(string = Current_Status,
                                                      pattern = "[Rr]ecords ([Ss]ent|[Ee]mail|[Pp]rovided)|[Ii]nformation [Ss]ent|[Rr]esponse[s]{0,1} [Ss]ent|reviewed|reside|responded|fulfilled") ~ "Sent",
                                           str_detect(string = Current_Status,
                                                      pattern = "[Pp]artial") ~ "Partial",
                                           str_detect(string = Current_Status,
                                                      pattern = "[Pp]ossession|no responsive") ~ "Not in possession",
                                           str_detect(string = Current_Status,
                                                      pattern = "[Nn][Oo][Tt]") ~ "Not sent",
                                           str_detect(string = Current_Status,
                                                      pattern = "^Open") ~ "Open",
                                           str_detect(string = Current_Status,
                                                      pattern = "[Ww]ithdrawn|withdrew") ~ "Withdrawn",
                                           str_detect(string = Current_Status,
                                                      pattern = "[Aa]gency|[Ww]eb|internet|DSCYTF|online|DHR|DOL") ~ "Referred elsewhere",
                                           str_detect(string = Current_Status,
                                                      pattern = "[Nn]o res[po]{2}nse") ~ "No response from requestor"))

foia_log_OMB %>% 
  count(Current_Status_Overall) %>% 
  arrange(-n)

foia_log_OMB %>% 
  count(Current_Status_Detail) %>% 
  arrange(-n)

```

Last, we reorganize our fields to make the data frame easier to understand.

```{r}

foia_log_OMB <- 
  foia_log_OMB %>% 
  select(Date_Received, 
         Requesting_Party, 
         Date_Assigned, 
         Assigned_to, 
         Division, 
         Information_Requested, 
         Date_Review, 
         Is_Noncustodial_Record, 
         Noncustodial_Provided_Other_Agency, 
         Internally_Reviewed_By, 
         Interim_communication, 
         Date_Response, 
         Date_Estimate_Sent, 
         All_Requested_Records_Sent, 
         All_Requested_Records_Sent_Flag, 
         All_Requested_Records_Sent_Detail, 
         Date_Requested_Records_Not_Provided, 
         Copying_Fees, 
         Administrative_Fees, 
         Date_of_Final_Disposition, 
         Current_Status, 
         Current_Status_Overall, 
         Current_Status_Detail)

```

```{r}

# Write to a .csv in the Data_Transformations folder
write_csv(foia_log_OMB, here("Data_Transformations", "FOIA_log_OMB_clean.csv"))

```

