---
title: "Visualizing OMB FOIA Data"
author: "Ryan Harrington"
output: html_document
---

# Overview

The Freedom of Information Act (FOIA) allows constituents to gain access to the data that their government creates. Gaining access to this data comes through a process called a "FOIA request". Making a FOIA request does not guarantee that the request will be fulfilled. There are many reasons that a request may be denied - from the data being unavailable to the data falling under executive privilege.

In Delaware, every government agency is required by law to keep a FOIA log. This log is a historical record of all FOIA requests that have been made to that agency. In an effort to better understand what data was commonly being FOIA requested, Open Data Delaware made a FOIA request to as many agencies as possible for their FOIA logs. Most of these FOIA-log-FOIA-requests were made on February 9, 2019 by David Ginzberg.

As the FOIA-log-FOIA-requests have been fulfilled, we have the opportunity to analyze each agency's FOIA logs to better understand their data. To begin this analysis, we look at the Office of Management and Budget's FOIA logs.

The data being used for this series of visualizations is not being presented in its raw form. The original data comes from the file `OMB FOIA TRACKING 032719.xlsx`. The data was processed and cleaned in the file `OMB_FOIA_Log_Cleaning.Rmd`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}

# For file management
# install.packages("here")
library(here)

# For data manipulation, functional programming, and graphing
# install.packages("tidyverse")
library(tidyverse)

# Loads the core tidyverse packages:
# - ggplot2, for data visualisation.
# - dplyr, for data manipulation.
# - tidyr, for data tidying.
# - readr, for data import.
# - purrr, for functional programming.
# - tibble, for tibbles, a modern re-imagining of data frames.
# - stringr, for strings.
# - forcats, for factors

# For interactive plots
# install.packages("plotly")
library(plotly)

```

# Import Data

```{r}

# Read in the `OMB FOIA TRACKING 032719.xlsx` file from the `Data` folder in the
# project directory and then store it in an object called foia_log_OMB
foia_log_OMB <- read_csv(here("Data_Transformations", "FOIA_log_OMB_clean.csv"))

```

The dataset consists of 23 fields. To the best of our ability, here is how those fields can be defined:

* `Date_Received` - the date that the FOIA request was received by OMB
* `Requesting_Party` - who made the FOIA request
* `Date_Assigned` - the date the the FOIA request was assigned to a person at OMB
* `Assigned_to` - who the FOIA request was assigned to at OMB
* `Division` - the division at OMB that the FOIA request was directed to
* `Information_Requested` - the information that the FOIA request is looking for
* `Date_Review` - date that the FOIA request was reviewed by OMB
* `Is_Noncustodial_Record` - whether or not the information sought after in the FOIA request is something that OMB has access to
* `Noncustodial_Provided_Other_Agency` - in the case that there is a noncustodial record, which agency would be more appropriate to review the request
* `Internally_Reviewed_By` - who the FOIA request will be reviewed by at OMB
* `Interim_communication` - any communication that has occurred between the date received and the date of final disposition
* `Date_Response` - date that the FOIA request was responded to
* `Date_Estimate_Sent` - date that an estimate for the cost of the FOIA request was sent
* `All_Requested_Records_Sent` - whether all requested records were sent to the requestor (original field)
* `All_Requested_Records_Sent_Flag` - whether requested records were sent to the requestor
* `All_Requested_Records_Sent_Detail` - detail regarding whether requested records were sent to the requestor
* `Date_Requested_Records_Not_Provided` - date or note regarding whether or not records were provided
* `Copying_Fees` - any copying fees (first 20 pages are free)
* `Administrative_Fees` - any administrative fees
* `Date_of_Final_Disposition` - date that the final records are sent
* `Current_Status` - status of the FOIA request (original)
* `Current_Status_Overall` - simplified status of the FOIA request
* `Current_Status_Detail` - simplified detail of the FOIA request

```{r}

foia_log_OMB

```


# Visualize Data

## Bar Charts

Who is assigned FOIA requests most frequently?
_Hint:_ Use the `Assigned_to` field

How can we make the background more attractive?
How can we change the fill color of the bars?

```{r}

foia_log_OMB %>% 
  ggplot(aes(x = Assigned_to)) +
  geom_bar()

```


```{r}

foia_log_OMB %>% 
  ggplot(aes(x = Assigned_to)) +
  geom_bar(fill = "red") +
  theme_minimal()

```

Which division receives the most FOIA requests?
_Hint:_ Use the `Division` field

How can we make the axis text more legible? Currently, many of the Divisions overlap with each other.

```{r}

foia_log_OMB %>%
  ggplot(aes(x = Division)) +
  geom_bar(fill = "red") +
  coord_flip() +
  theme_minimal()

foia_log_OMB %>%
  ggplot(aes(x = Division)) +
  geom_bar(fill = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1))

```

What is the current status of all FOIA requests ever filed?
_Hint:_ Try the fields `Current_Status_Overall` and `Current_Status_Detail`. Which do you prefer?

Can we change the labels of our fields?

```{r}

foia_log_OMB %>% 
  ggplot(aes(x = Current_Status_Overall)) +
  geom_bar() +
  coord_flip() +
  theme_minimal() +
  labs(x = "Current Status Overall",
       y = "Count")

foia_log_OMB %>% 
  ggplot(aes(x = Current_Status_Detail)) +
  geom_bar() +
  coord_flip() +
  theme_minimal() +
  labs(x = "Current Status Detail",
       y = "Count")

```

## Line Charts

Are there any months with particularly high numbers of FOIA requests?

How can we add a title to our graph?
How else can we manipulate the aesthetics of our graph?

```{r}

foia_log_OMB %>%
  mutate(Month_Received = as.Date(format(Date_Received, format = "%Y-%m-01"))) %>% 
  count(Month_Received) %>% 
  ggplot(aes(x = Month_Received,
             y = n)) +
  geom_line(color = "red",
            size = 1) +
  theme_minimal() +
  labs(x = "Month FOIA Request was Made",
       y = "Count of FOIA Requests",
       title = "Number of FOIA Requests by Month")

```

Does the current status of the FOIA request change based upon when it was filed?

How can we use multiple geometries to create more complex graphs?
How can we improve our theme?

```{r}

foia_log_OMB %>%
  mutate(Month_Received = as.Date(format(Date_Received, format = "%Y-%m-01"))) %>% 
  count(Month_Received, Current_Status_Overall) %>% 
  group_by(Current_Status_Overall) %>% 
  complete(Month_Received = seq.Date(min(Month_Received), 
                                     max(Month_Received), by = "months")) %>% 
  replace(is.na(.), 0) %>% 
  ggplot(aes(x = Month_Received,
             y = n,
             group = Current_Status_Overall,
             color = Current_Status_Overall)) +
  geom_line() + 
  geom_point(size = 1) +
  labs(x = "Month FOIA Request was Made",
       y = "Count of FOIA Requests",
       title = "Number of FOIA Requests by Month",
       subtitle = "By Current Status") +
  theme_minimal() +
  theme(axis.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"))

```

Are particular divisions more likely to receive a FOIA request at a given time of the year?

```{r}

foia_log_OMB %>%
  mutate(Month_Received = as.Date(format(Date_Received, format = "%Y-%m-01"))) %>% 
  count(Month_Received, Division) %>% 
  group_by(Division) %>% 
  complete(Month_Received = seq.Date(min(Month_Received), 
                                     max(Month_Received), by = "months")) %>% 
  mutate(n = replace_na(n, 0)) %>% 
  ungroup() %>% 
  ggplot(aes(x = Month_Received,
             y = n,
             group = Division,
             color = Division)) +
  geom_line() +
  labs(x = "Month FOIA Request was Made",
       y = "Count of FOIA Requests",
       title = "Number of FOIA Requests by Month",
       subtitle = "By Division") +
  theme_minimal() +
  theme(axis.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"))

```

What might be a better way to graph this?

```{r}

foia_log_OMB %>%
  mutate(Month_Received = as.Date(format(Date_Received, format = "%Y-%m-01"))) %>% 
  count(Month_Received, Division) %>% 
  group_by(Division) %>% 
  complete(Month_Received = seq.Date(min(Month_Received), 
                                     max(Month_Received), by = "months")) %>% 
  mutate(n = replace_na(n, 0)) %>% 
  ungroup() %>% 
  ggplot(aes(x = Month_Received,
             y = n)) +
  geom_line() +
  facet_wrap(~Division, nrow = 5) +
  labs(x = "Month FOIA Request was Made",
       y = "Count of FOIA Requests",
       title = "Number of FOIA Requests by Month",
       subtitle = "By Division") +
  theme_minimal() +
  theme(axis.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"))

```

## Bonus: Interactivity

How can we make our plots interactive?

```{r}

ggplotly(
  
  foia_log_OMB %>%
  mutate(Month_Received = as.Date(format(Date_Received, format = "%Y-%m-01"))) %>% 
  count(Month_Received, Division) %>% 
  group_by(Division) %>% 
  complete(Month_Received = seq.Date(min(Month_Received), 
                                     max(Month_Received), by = "months")) %>% 
  mutate(n = replace_na(n, 0)) %>% 
  ungroup() %>% 
  ggplot(aes(x = Month_Received,
             y = n)) +
  geom_line() +
  facet_wrap(~Division, nrow = 5) +
  labs(x = "Month FOIA Request was Made",
       y = "Count of FOIA Requests",
       title = "Number of FOIA Requests by Month",
       subtitle = "By Current Status") +
  theme_minimal() +
  theme(axis.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"))
  
)

```

